
\begin{acknowledgments}

The researchers would like to express their deepest gratitude to the individuals whose guidance, support, and encouragement made this thesis possible.

First and foremost, we extend our sincere appreciation to Rosel O. Onesa, our Dean and Thesis 1 instructor, for her unwavering guidance, insightful feedback, and dedication in strengthening our foundation as researchers. Her mentorship during the initial stages of this study greatly contributed to the direction and structure of our work.

Our profound thanks also go to Ms. Tiffany Lyn O. Pandes,  our Thesis 2 instructor and respected panel member, for her continuous support, constructive suggestions, and patience throughout the development and completion of this project.

We would like to express our heartfelt gratitude to Ms. Kaela Marie Fortuno,  our Adviser, whose expertise, encouragement, and professional guidance were vital to the success of this study. Her consistent support and commitment inspired us to pursue excellence.

Special thanks are also extended to Sir Allan Ibo, our Consultant, for sharing his technical knowledge, especially in areas requiring advanced expertise. His contributions significantly enhanced the technical reliability of our system.

We gratefully recognize Ms. Ma. Allaine Agna our Grammarian, for meticulously reviewing our manuscript and ensuring clarity, coherence, and adherence to academic standards.

Our sincere appreciation is likewise extended to the Thesis Panel, chaired by Ian P. Benitez, with members Ms. Tiffany Lyn O. Pandes, and  Rosel O. Onesa, for their professional evaluation, constructive recommendations, and guidance that improved the overall quality of this research.

We would also like to acknowledge the riders who willingly participated in our data collection process. Their cooperation in allowing us to capture images and videos was essential for building and training our dataset, and without their assistance, the development of our thesis project would not have been possible.

Deep appreciation is given to the College of Computer Studies, along with all faculty members who have imparted knowledge and skills throughout our academic journey.

We extend heartfelt thanks to our families, friends, and loved ones for their continuous support, patience, and motivation throughout the challenging phases of this study. Their encouragement gave us strength to persevere.

Above all, we express our gratitude to God, whose guidance, wisdom, and blessings sustained us from the beginning until the completion of this thesis.
To everyone who contributed—whether directly or indirectly—to the success of this study, thank you very much.


\end{acknowledgments}
